Microxanox - an R package for simulating an $MICR$obial ecosystem that can occupy $OX$ic or $ANOX$ic states.

### 2022-05-03 inst/manuscript/manuscript_microxanox.Rmd
Update manuscript_microxanox.Rmd

# Motivation and significance
Many ecosystems are exposed to gradual changes of environmental variables, to which the responses are not always as gradual as the change of the environmental variable. One example where the gradual change of a single environmental causing an abrupt change of the system is the switch from an aerobic to anaerobic system. This system has been investigated by @Bush2017. We wanted to take this investigation one step further, and look at the role an increased biodiversity plays in these dynamics [@REF_NEEDED]. For this purpose, we developed this package.

@@

The ODEs for the rates of change are specified in the function `bushplus_dynamic_model()`. This augmented version of the model published in [@Bush2017] can handle multiple strains within each of the three functional groups, temporal variation in oxygen diffusivity, and events.

@@

1. all parameter needed to run a simulation or find a stable state are contained in a single parameter object. This object is created by using the functions `new_..._parameter()`, `new_initial_state()` and `new_strain_parameter()`. Which one of the `new_..._parameter()` functions has to be used when, will be discussed in the section  \@ref(runsim) and in more detail in the [User Guide](LINK_NEEDED).
2. The function call `run_...(parameter)` will run the simulation using the parameter as defined in the object `parameter`.
3. The return value of the `run_...(parameter)` function is identical to the parameter object plus an additional slot named `results` which contains the results of the run
4. As the return value contains all parameter, it is possible to re-run the simulatuion by simply running `run_...(result)`.

@@

## Finding a Steady State of the model

@@

### Replication Method
The replication method is implementad in the function `run_replication_ssfind()` which takes a parameter object as returned by the function `new_replication_ssfind_parameter()` and the number of cores for multithreading the simulation.

### Temporal Method
The temporal method involves two simulations for a particular system configuration (parameter set). In one simulation the oxygen diffusivity is *increased* in a step-wise fashion. In the other it is *decreased* in a step-wise fashion. That is, oxygen diffusivity is held at a constant level for long enough for steady state to be reach, that state is recorded, and then a slightly higher (or lower) oxygen diffusivity value is set. Hence, at that time point, the system is effectively started with initial conditions that are the state of the system in the previous time step.

@@

## Extracting Stability Measures
From the raw results returned by these `run_...()` functions, the stability measures can be extracted by using the function `get_stability_measures()`. These measures include non-linearity and hysteresis measures, of the response of the simulated system to environmental change.

## Use Cases
We will now show two use cases which illustrate applications of the package. All use cases can be used as starting points for other investigations of the behavior of the system.

@@

Ass all objects were saved, results can be easily reproduced and parameter can be changed to assess the impact of these changes (e.g. to conduc a sensitivity analysis).

@@

As discussed in the paper [@REF_NEEDED], the role biodiversity plays in abrupt regime shifts based on gradual changing environmental parameter is not well understood. This model (as part of the package) has been used to investigate these dynamics and the results are available in @REF_NEEDED.

@@

The free and open source implementation and extension of the model used in @Bush2017 provides the means of reproducing the results published while at the same time provides the means of doing investigations which extend the original publication, as done in @REFNEEDED.

>>>>

# Motivation and significance
Many ecosystems are exposed to gradual changes of environmental variables, to which the responses are not always as gradual as the change of the environmental variable. One example where the gradual change of a single environmental variable is causing an abrupt change of the system is the switch from an aerobic to anaerobic system. This system has been investigated by @Bush2017. We wanted to take this investigation one step further, and look at the role an increased biodiversity plays in these dynamics [@REF_NEEDED]. For this purpose, we developed this package.

@@

The model by @Bush2017 is based on ODEs for the rates of change which are specified in the function `bushplus_dynamic_model()`. This augmented version of the model published in [@Bush2017] can handle multiple strains within each of the three functional groups, temporal variation in oxygen diffusivity, and events.

@@

1. all parameter needed to run a simulation or find a stable state are contained in a single parameter object. This object is created by using the functions `new_..._parameter()`, `new_initial_state()` and `new_strain_parameter()`. Which one of the `new_..._parameter()` functions has to be used when, will be discussed in the section  \@ref(runsim) and in more detail in the [User Guide](LINK_NEEDED).
2. The function call `run_...(parameter)` will run the simulation using the parameter as defined in the object `parameter`.
3. The return value of the `run_...(parameter)` function is identical to the parameter object plus an additional slot named `results` which contains the results of the run
4. As the return value contains all parameter, it is possible to re-run the simulation by simply running `run_...(result)`.

@@

### Finding a Steady State of the model

@@

#### Replication Method
The replication method is implementad in the function `run_replication_ssfind()` which takes a parameter object as returned by the function `new_replication_ssfind_parameter()` and the number of cores for multithreading the simulation.

#### Temporal Method
The temporal method involves two simulations for a particular system configuration (parameter set). In one simulation the oxygen diffusivity is *increased* in a step-wise fashion. In the other it is *decreased* in a step-wise fashion. That is, oxygen diffusivity is held at a constant level for long enough for steady state to be reach, that state is recorded, and then a slightly higher (or lower) oxygen diffusivity value is set. Hence, at that time point, the system is effectively started with initial conditions that are the state of the system in the previous time step.

@@

### Extracting Stability Measures
From the raw results returned by these `run_...()` functions, the stability measures can be extracted by using the function `get_stability_measures()`. These measures include non-linearity and hysteresis measures, of the response of the simulated system to environmental change.

## Use Cases
We will now show three use cases which illustrate applications of the package. All use cases can be used as starting points for other investigations of the behavior of the system.

@@

As all objects were saved, results can be easily reproduced and parameter can be changed to assess the impact of these changes (e.g. to conduct a sensitivity analysis).

@@

As discussed in the paper [@REF_NEEDED], the role biodiversity plays in abrupt regime shifts based on gradual changing environmental parameter is not well understood. This model (as part of the package) has been used to investigate these dynamics and the results are available in with the DOI DOI NEEDED @REF_NEEDED.

@@

The free and open source implementation and extension of the model used in @Bush2017 provides the means of reproducing the results published while at the same time provides the means of doing investigations which extend the original publication, as done in @REF_NEEDED.
